Temporal ordering and registration of imaging data in cross--sectional studies of development

### 2014-03-27 image_analysis_paper/pnas_draft.tex
updated paper

Note that both of these tasks have been accomplished based on prior knowledge about the underlying developmental processes: We knew that the peak of the green signal can be used to align the images and that the progress of cellularization can order them in time. 
%
Is this information essential? 
%
Below we show that this is not the case and that both the temporal ordering and registration of images can be fully automated, without prior information about the underlying dynamics. 
%
Our algorithmic approach is based on diffusion maps, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets. 
%
The paper is organized as follows. 
%
In the next two sections, we use a simpler version of the data (one-dimensional spatial profiles) to formulate both tasks mathematically.  
%
Then, we show that both of the tasks can be effectively accomplished in single step. 
%
Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align the images and order them correctly in time. 
%
We conclude by discussion of the general requirements that must be satisfied for our approach to work and compare it to other ideas in cross-sectional studies of biological dynamics.

@@

We will first demonstrate our techniques considering {\em only} considering rotations. 
%
We have converted the two-dimensional images shown in Figure \ref{subfig:images_scrambled} into concentration profiles on a ring (we will address the full two-dimensional images in a subsequent section).
%
Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional concentration profile; the profile is discretized into 100 points. 
%
Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one image.

@@

Often when one collects images in experiments, each image is taken in a different orientation. 
%
Therefore, the first task in image analysis is to align or {\em register} the set of images. 
%
This requires translating and rotating each image, so that the set of images are globally aligned.

There are many methods for image registration. 
%
``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
%
However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
Instead, we will use a relatively new technique called angular synchronization\cite{singer2011angular} to register sets of images.
%
Angular synchronization first computes the pairwise alignments (translation + rotation) for every possible pair of images, effectively using each image as a template for every other image.
%
From this large collection of pairwise alignments, it then tries to find the optimal translation and rotation for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
%
The algorithm solves this optimization problem by relaxing it to an eigenproblem; the top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the translation and rotation for each image.
%
Because the method considers {\em all} pairwise alignments, angular synchronization is often much more robust to noise.

Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization;
the peaks are aligned, even though we input no information about the existence or location of these peaks into our algorithm.
%
We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.

@@

We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
Although our data are high-dimensional (for example, we will work with 2-color images of 10000 pixels, and so we treat these images as 20000-dimensional vectors, where each dimension describes the intensity of one color in one pixel), we will assume that they lie on a one-dimensional, perhaps nonlinear curve.
%
We further assume that curve can be parameterized by time.
%
Therefore, our goal is to {\em automatically} uncover a parameterization of this curve, so that we can sort/order our data along it, effectively ordering our data in time.

To ``unravel'' this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve.
%
We want to find a parameterization of the points, such that the points which are close in the original close are {\em as close as possible} in the parameterization.
%
This problem can be formulated as an optimization problem \cite{...}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points.

As an illustration of how local neighbors are important for parameterizing a curve, Figure~\ref{subfig:dmaps_edges} shows a data set which lies on a one-dimensional, nonlinear curve in two dimensions.
%
The edges between the pairs of data points are colored according to the distance between the data points, so that close data points have a dark edge between them.
%
Visually, these dark edges are meaningful and informative of the structure of the data; they clearly delineate the one-dimensional curve in the data that our eye can extract.
%
Furthermore, the light edges are not informative and do not give any meaningful structure to the curve. 
%
The DMAPS parameterization of these are data are shown in Figure~\ref{subfig:dmaps_color}.
%
The parameterization is what we would expect visually.
%
Although this data is useful for illustrative purposes, it is

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
%
We use the total Euclidean distance (both the red and the green channels) between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) DMAPS component, $\phi_2$. 
%
As expected, the intensity of the dpERK peaks grows as a function of time. 
%
Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
%
Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.

@@

In many cases, we are interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
%
It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.

Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
%
The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) VDM coordinate.

@@

The same methodologies (with small modifications) can be applied to two-dimensional images.
%
Instead of having $100$-dimensional concentration vectors, we now work with images containing $10000$ pixels.
%
The two-dimensional images are registered with respect to all possible translations and rotations; 
we convert two-dimensional translations and rotations to three-dimensional rotations so that we can apply angular synchronization or vector diffusion maps. 
%
We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_scrambled}.
%
The results are shown in Figure~\ref{subfig:images_vdm}.
%
The peaks are now aligned, and the intensity of dpERK grows as a function of time, as expected.

In these {\em Drosophila} experiments, we have the luxury of an independent time marker from membranes.
%
We can therefore evaluate the quality of our orderings obtained from VDM by comparing with the orderings obtained from the membrane markers.
%
The rank of each data point, computed using VDM, versus the rank of each data point computed from the membrane thickness, is shown in Figure~\ref{subfig:rank_corr}.
%
The rank correlation is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from the membrane thickness.
%
There is more scatter for the younger embryos; this is to be expected, as the dpERK signal is very low in this range.

Although our methods are fairly robust, they do require a sufficient amount of data to ``see'' a smooth developmental trajectory.
%
We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering using the membrane markers.
%
The results are shown in Figure~\ref{subfig:bootstrap}.
%
For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate. 
%
However, for moderately sized data sets (approximately 30 data points), the 
rank correlation begins to plateau and VDM can accurately order the data.

@@

We have presented mathematical techniques for registration and temporal ordering of images arising in studies of developmental biology.
%
Although we illustrated our methods using a specific data set arising in a study of {\em Drosophila} embryogenesis, the techniques are very general and require little knowledge of the image features or the developmental dynamics.

We would like to note that there have been multiple other works with temporally ordering biological data (RNA data, cell morphology data, disease progression data, etc.). 
%
The majority of these works temporally order the data either by solving a travelling salesman probelem, or by constructing a minimum spanning tree on the data.
%
We, instead, use nonlinear dimensionality reduction techniques, to construct a one-dimensional embedding of the data.
%
Furthermore, we work with images that must be first registered before they can be temporally ordered.

Although our methods are applicable to a wide variety of data sets, there are several questions for which our techniques can provide no answers.
%
Because our registration algorithm only considers pairwise information, we cannot recover the true ``zero'' point of the images. 
%
Similairly, we cannot know which end of our trajectory is the beginning and which is the end.
%
Both of these tasks bust be done by hand as a post-processing step.
%
Furthermore, our methods can only provide a temporal ordering for the data, and cannot provide any information as to the speed or absolute time of the data samples.
%
Regardless, we are confident that our methods hold great promise in areas where cross-sectional imaging data is readily available and relatively inexpensive to obtain. 
%
There are many applications for which develpomental dynamics are of interest, and yet there is no time marker to use to reconstruct the dynamics from such cross-sectional data.

@@

\section{Angular synchronization}

@@

\section{Diffusion Maps}

@@

We then want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
%
We therefore want to solve the following optimization problem
\begin{equation} \label{eq:dmaps_opt_problem}
\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))
\end{equation}
%
%The solution \ref{eq:dmaps_opt_problem} is $\phi_2$, the second eigenvector of $A$, where $A = D^{-1} W$, and $D$ is a diagonal matrix with $D_{ii} = \sum_j W_{ij}$ \cite{...}.
%%
%Therefore, $\phi_2$ provides a one-dimensional parameterization of the data, and ordering the data points by the value of $\phi_2$ will order them in time.

@@

\section{Vector Diffusion Maps}

@@

We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
%
Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image, and the optimal rotations can be computed as described for angular synchronization.
%
However, the eigenvectors now also contain information about the embedding coordinates for our images.
%
In general, the embedding coordinates are given by 
\begin{equation}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
%
If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e. the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
%
If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, in order to uncover the true manifold modulo symmetries.

@@

Registering images with respect to rotations {\em and translations} requires some additional care.
%
%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations. 
%
For \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
%
However, the typical matrix representation for the group of two-dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property. 
%
Instead, we can (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices.
%
We do this by projecting the image onto a portion of the surface of a (three-dimensional) sphere (see Figure \ref{fig:SO3_picture}).
%
Rotations in two-dimensions now correspond to rotations around one principal axis in three-dimensions, and translations in two-dimensions correspond (approximately) to rotations around the other two principal axes in three-dimensions. (Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
%
These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.

@@

From the three-dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
we build the matrix $H$ in \eqref{eq:H_to_R} and compute the estimates $R_{i, est}$ in \eqref{eq:R_est}.
%
From the matrices $R_{1, est}, \dots, R_{m, est}$, we now wish to find the corresponding translations and rotations of the images.
%
Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
\begin{equation}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) \\
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) \\
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta) 
\end{aligned}
\end{equation}
%
We then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
%
We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.  
\end{materials}

@@

\begin{figure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{fish1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{fish2}
\caption{}
\end{subfigure}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now registered and temporally ordered. This is easy to do by hand because the data set is small and the developmental changes are easy to recognize.} 
\label{fig:fish}
\end{figure}

\begin{figure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{fig2a}
\caption{}
\label{subfig:images_scrambled}
\end{subfigure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{fig2b}
\caption{}
\label{subfig:images_hand_ordered}
\end{subfigure}
\caption{{\em Drosophila} data. (a) Dorsal-ventral images of {\em Drosphila} embryos in nuclear cycle 14, each stained for Dorsal (green) and dpERK (red). (b) Images from (a), now registered using the location of the Dorsal peak, and temporally ordered using the membrane marker.} 
\label{fig:fluorescent_images}
\end{figure}

>>>>

Note that both of these tasks have been accomplished based on prior knowledge about the underlying developmental processes: We knew that the peak of the green signal can be used to align the images and that the progress of cellularization can order them in time. 
%
Is this information essential? 
%
Below we show that this is not the case and that both the temporal ordering and registration of images can be fully automated, without prior information about the underlying dynamics. 
%
Our temporal ordering analysis is based on diffusion maps, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets.
%
Our image registration is done via angular synchronization, a relatively new technique for registering images which has been applied to reconstructing the three-dimensional structure of molecules from cryo-electron microscopy \cite{singer2011three}; here, we will use the algorithm to register {\em Drosophila} images.
%
The paper is organized as follows. 
%
In the next two sections, we use a simpler version of the data (one-dimensional spatial profiles) to formulate both tasks mathematically.  
%
Then, we show that both of the tasks can be effectively accomplished in single step. 
%
Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align the images and order them correctly in time. 
%
We conclude by discussion of the general requirements that must be satisfied for our approach to work and compare it to other ideas in cross-sectional studies of biological dynamics.

@@

We will first demonstrate our techniques using one-dimensional concentrations profiles.
%
We have converted the two-dimensional images shown in Figure \ref{subfig:images_scrambled} into profiles defined on a ring, corresponding to the concentrations around the periphery of the embryo (we will address the full two-dimensional images in a subsequent section).
%
Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional profile; the profile is discretized into 100 points, and so our signals will be 200-dimensional vectors (100 dimensions correspond to the red channel at the 100 discretization points, and 100 dimensions correspond to the green channel at the 100 discretization points). 
%
Rotation of this image corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure \ref{subfig:1d_image_example}. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one image.
%
The task is now to shift each profile left and right to optimally register/align the signals, and then temporally order the rows in the array to reconstruct the developmental trajectory.

@@

Often when one collects images in experiments, each image is taken in a different orientation. 
%
Therefore, the first task in image analysis is to align or {\em register} the set of images. 
%
This requires rotating and/or translating each image, so that the set of images are globally aligned.

There are many methods for image registration. 
%
``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
%
However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
Instead, we will use a relatively new technique called angular synchronization\cite{singer2011angular} to register sets of images.
%
Angular synchronization first computes the pairwise alignments (rotation for our one-dimensional profiles in Figure \ref{subfig:1d_unaligned_unordered}, or translation + rotation for the two-dimensional images in Figure \ref{subfig:images_scrambled}) for every possible pair of images, effectively using each image as a template for every other image.
%
From this large collection of pairwise alignments, it then tries to find the optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
%
The algorithm solves this optimization problem by relaxing it to an eigenproblem; the top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignment for each image.
%
Because the method considers {\em all} pairwise alignments, angular synchronization is often much more robust to noise.

Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization;
the peaks are aligned, even though we input no information about the existence or location of these peaks into our algorithm.
%
We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation; we are permitted rotate every point by a constant rotation.

@@

We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
Although our data are high-dimensional (for example, we will work with 2-color images of 10000 pixels, and so our data are 20000-dimensional vectors, where each dimension describes the intensity of one color in one pixel), we will assume that they lie on a one-dimensional, but perhaps nonlinear, curve.
%
We further assume that this curve is monotonic in time.
%
Therefore, our goal is to {\em automatically} uncover a parameterization of this curve, so that we can sort/order our data along it, effectively ordering our data in time.
%
To parameterize this one-dimensional curve which lies in high-dimensional space, we assume that data points which are close in the high-dimensional space should also be close on the curve, 
e.g, images which look similar will be close in the high-dimensional pixel space (they will have similar pixel intensities), and we would like these similar images to be close to each other in our trajectory.
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.

As an illustration of how local neighbors are important for parameterizing a curve, Figure~\ref{subfig:dmaps_edges} shows a data set which lies on a one-dimensional, nonlinear curve in two dimensions.
%
The edges between the pairs of data points are colored according to the distance between the data points, so that close data points have a dark edge between them.
%
Visually, these dark edges are meaningful and informative of the structure of the data; they clearly delineate the one-dimensional curve in the data that our eye can extract.
%
Furthermore, the light edges are not informative and do not give any meaningful structure to the curve.

The problem of finding a parameterization of the data which forces nearby points together can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points. 
%
This procedure is called diffusion maps \cite{coifman2005geometric}; the formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization). 
%
The DMAPS parameterization of these are data are shown in Figure~\ref{subfig:dmaps_color}.
%
The parameterization agrees with what our eye can detect visually.
%
However, we will work with very high-dimensional data, rather than these two-dimensional data points; we therefore have no easy way to plot and visualize the data, and rather need DMAPS to automatically uncover such structure in the data.

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we would like to use DMAPS to order them in time.
%
We use the total Euclidean distance (both the red and the green channels) between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) DMAPS component, $\phi_2$. 
%
As expected, the intensity of the dpERK peaks grows as a function of time. 
%
Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
%
Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.

@@

In many cases, we are interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using VDM. 
%
The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) VDM coordinate.

@@

The same methodologies (with small modifications) can be applied to two-dimensional images.
%
Instead of having $200$-dimensional concentration vectors, we now work with $20000$-dimensional images ($100x100$ pixel images containing two color channels).
%
The two-dimensional images are registered with respect to translations and rotations 
(we convert the two-dimensional translations and rotations to three-dimensional rotations so that we can apply angular synchronization or vector diffusion maps). 
%
We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_scrambled}.
%
The results are shown in Figure~\ref{subfig:images_vdm}.
%
The peaks are now aligned, and the intensity of dpERK grows as a function of time, as expected.

In these {\em Drosophila} experiments, we have the luxury of an independent measure of time from cellularization.
%
We can therefore evaluate the quality of our orderings obtained from VDM by comparing with the orderings obtained from cellularization.
%
The rank of each data point, computed using VDM, versus the rank of each data point computed from cellularization, is shown in Figure~\ref{subfig:rank_corr}.
%
The rank correlation is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from cellularization.
%
There is more scatter for the younger embryos; this is to be expected, as the dpERK signal, which is the main feature that changes with time, is very low in this range.

Although our methods are fairly robust, they do require a sufficient amount of data so that DMAPS can ``see'' a smooth trajectory.
%
We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering from cellularization.
%
The results are shown in Figure~\ref{subfig:bootstrap} (each data point is averaged over 50 random subsamples).
%
For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate. 
%
However, for moderately sized data sets (approximately 30 data points), the 
rank correlation begins to plateau and VDM can accurately order the data.

@@

We have presented mathematical techniques for registration and temporal ordering of images arising in studies of developmental biology.
%
Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, the techniques are very general and require little knowledge of the image features or the developmental dynamics.

We would like to note that there have been multiple other works with temporally ordering biological data (RNA-seq, cell morphology, disease progression, etc.). 
%
The majority of these works temporally order the data either by solving a traveling salesman problem, or by constructing a minimum spanning tree on the data.
%
We, instead, use nonlinear dimensionality reduction techniques, to construct a one-dimensional embedding of the data.
%
Furthermore, we work with images that must be first registered before they can be temporally ordered.

Although our methods are applicable to a wide variety of data sets, there are several questions for which our techniques can provide no answers.
%
Because our registration algorithm only considers pairwise information, we cannot recover the true ``zero'' point of the images. 
%
Similairly, we cannot know which end of our trajectory is the beginning and which is the end.
%
Both of these tasks bust be done by hand as a post-processing step.
%
Furthermore, our methods can only provide a temporal ordering for the data, and cannot provide any information as to the speed or absolute time of the data samples.
%
Regardless, we are confident that our methods hold great promise in areas where cross-sectional imaging data is readily available and relatively inexpensive to obtain. 
%
They provide a simple and cost-effective alternative to live imaging or ordering images by hand in order to gain insight into the developmental dynamics.

@@

\section{Angular synchronization \cite{singer2011angular}}

@@

\section{Diffusion Maps \cite{coifman2005geometric}}

@@

We then want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
%
We therefore want to solve the following optimization problem \cite{Belkin2003}
\begin{equation} \label{eq:dmaps_opt_problem}
\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))
\end{equation}
%
%The solution \ref{eq:dmaps_opt_problem} is $\phi_2$, the second eigenvector of $A$, where $A = D^{-1} W$, and $D$ is a diagonal matrix with $D_{ii} = \sum_j W_{ij}$ \cite{...}.
%%
%Therefore, $\phi_2$ provides a one-dimensional parameterization of the data, and ordering the data points by the value of $\phi_2$ will order them in time.

@@

\section{Vector Diffusion Maps \cite{singer2012vector}}

@@

We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
%
Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image, and the optimal rotations can be computed as described for angular synchronization.
%
However, the eigenvectors now also contain information about the embedding coordinates for our images.
%
In general, the embedding coordinates are given by 
\begin{equation}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
%
If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e., the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
%
If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, in order to uncover the true manifold modulo symmetries.

@@

Registering images with respect to rotations {\em and translations} requires some additional care.
%
%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations. 
%
For \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
%
However, the typical matrix representation for the group of two-dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property. 
%
Instead, we can (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices.
%
We do this by projecting the image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}. %(see Figure \ref{fig:SO3_picture}).
%
Rotations in two-dimensions now correspond to rotations around one principal axis in three-dimensions, and translations in two-dimensions correspond (approximately) to rotations around the other two principal axes in three-dimensions. (Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
%
These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.

@@

From the three-dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
we build the matrix $H$ in \eqref{eq:H_to_R} and compute the estimates $R_{i, est}$ in \eqref{eq:R_est}.
%
From the matrices $R_{1, est}, \dots, R_{m, est}$, we now wish to find the corresponding translations and rotations of the images.
%
Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
\begin{equation}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) \\
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) \\
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta) 
\end{aligned}
\end{equation}
%
We then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
%
We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.

\section{Rank correlation}

We use Spearman's rank correlation coefficient as a metric to evaluate the quality of our orderings.
%
Given data $x_1, \dots, x_m$ and $y_1, \dots, y_m$ for which we would like to compute the rank correlation coefficient, we first compute the ranks $X_i$ and $Y_i$ of each data point $x_i$ and $y_i$ in their respective data sets.
%
The rank correlation coefficient, $\rho$, is then given by
\begin{equation}
\rho = \frac{\sum_{i=1}^m (X_i - \overline{X})(Y_i - \overline{Y})}{\sqrt{\sum_{i=1}^m (X_i - \overline{X})^2 \sum_{i=1}^m (Y_i - \overline{Y})^2}}
\end{equation}
where $\overline{X}$ and $\overline{Y}$ are the means of $X_1, \dots, X_m$ and $Y_1, \dots, Y_m$, respectively.

\end{materials}

@@

\begin{figure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{fish1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{fish2}
\caption{}
\end{subfigure}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now registered and temporally ordered. For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.} 
\label{fig:fish}
\end{figure}

\begin{figure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{fig2a}
\caption{}
\label{subfig:images_scrambled}
\end{subfigure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{fig2b}
\caption{}
\label{subfig:images_hand_ordered}
\end{subfigure}
\caption{{\em Drosophila} data. (a) Dorsal-ventral images of {\em Drosphila} embryos, each fluorescentally stained for Dorsal (green) and dpERK (red). (b) Images from (a), now registered using the location of the Dorsal peak, and temporally ordered using cellularization.} 
\label{fig:fluorescent_images}
\end{figure}

\begin{figure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{dmaps_schematic_edges}
\caption{}
\label{subfig:dmaps_edges}
\end{subfigure}
\begin{subfigure}{0.25\textwidth}
\includegraphics[width=\textwidth]{dmaps_schematic_color}
\caption{}
\label{subfig:dmaps_color}
\end{subfigure}
\caption{Diffusion maps illustration. (a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges). (b) The data in (a), colored by the first (non-trivial) DMAPS eigenvector, $\phi_2$. The color parameterizes the one-dimensional curve. }
\label{fig:dmaps_schematic}
\end{figure}
